The all-seeing eye of resonant Auger electron spectroscopy: a study on aqueous KCl

### 2018-04-25 abstract.tex
apr 25
jpcl draft

X-ray absorption and Auger electron spectroscopies are powerful tools to probe the electronic structure and immediate surroundings of ions in solution. In this work we use a combination of these methods together with {\it ab initio} calculations to study the electronic structure and decay in aqueous KCl at the K-edges of \ki~and \cli. Although the two ions are isoelectronic, their Auger electron spectra exhibit notably different features. %To explain these differences, we carried out {\it ab initio} calculations of both the core excited states and the final Auger states of bare \ki, \cli~and their microsolvated clusters. 
This is due to the fact that the energetic order of the 1s$^{-1}$3d and 1s$^{-1}$4p core excited states is inverted in \ki~with respect to \cli, and moreover the two states lie energetically close in the bare \ki~ion leading to their mixing in the presence of the solvent. As a result, the dipole forbidden 1s$^{-1}$3d state is populated upon K-shell excitation of aqueous K$^{+}$. The resonant Auger decay of this state results in a separate feature in the Auger electron spectrum of \ki~which is absent in the spectrum of \cli. The results of this work represent a pioneering study of the decay processes initiated by photoabsorption in the tender x-ray regime close to threshold in liquids and are thus of importance in unveiling the mechanisms of radiation damage in biologically relevant systems.
%Denis: line 2, I replaced spectroscopies by methods

>>>>

X-ray absorption and Auger electron spectroscopies are powerful tools to probe the electronic structure and immediate surroundings of ions in solution. Using these techniques, we study the electronic structure and decay of aqueous KCl at the K-edges of \ki~and \cli. Although the two ions are isoelectronic, their Auger electron spectra exhibit notably different features. This is due to the energetic proximity of the 1s$^{-1}$3d and 1s$^{-1}$4p core excited states in the bare \ki~ion leading to their mixing in the presence of the solvent. As a result, the dipole forbidden 1s$^{-1}$3d state is populated upon K-shell excitation of aqueous K$^{+}$, and its resonant Auger decay of this state results in a separate feature in the Auger electron spectrum of \ki~which is absent in the spectrum of \cli. The results of this work represent a pioneering study of the decay processes initiated by photoabsorption in the tender x-ray regime close to threshold in liquids.

%X-ray absorption and Auger electron spectroscopies are powerful tools to probe the electronic structure and immediate surroundings of ions in solution. In this work we use a combination of these methods together with {\it ab initio} calculations to study the electronic structure and decay in aqueous KCl at the K-edges of \ki~and \cli. Although the two ions are isoelectronic, their Auger electron spectra exhibit notably different features. To explain these differences, we carried out {\it ab initio} calculations of both the core excited states and the final Auger states of bare \ki, \cli~and their microsolvated clusters. This is due to the fact that the energetic order of the 1s$^{-1}$3d and 1s$^{-1}$4p core excited states is inverted in \ki~with respect to \cli, and moreover the two states lie energetically close in the bare \ki~ion leading to their mixing in the presence of the solvent. As a result, the dipole forbidden 1s$^{-1}$3d state is populated upon K-shell excitation of aqueous K$^{+}$. The resonant Auger decay of this state results in a separate feature in the Auger electron spectrum of \ki~which is absent in the spectrum of \cli. The results of this work represent a pioneering study of the decay processes initiated by photoabsorption in the tender x-ray regime close to threshold in liquids and are thus of importance in unveiling the mechanisms of radiation damage in biologically relevant systems.
%Denis: line 2, I replaced spectroscopies by methods

### 2018-04-25 introduction.tex
apr 25
jpcl draft

It has been well understood in atoms and molecules in gas phase that the course of a decay cascade depends on the character of the initially populated states \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415,Miteva14:164303,travnikova16:213001,Gokhberg14:661,Trinter14:664}.
%The course of a decay cascade depends on the character of the initially populated states. This has been well understood in atoms and molecules in gas phase  \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415,Miteva14:164303,travnikova16:213001,Gokhberg14:661,Trinter14:664}. In the case of a core ionized state, the Auger decay process, designated as normal Auger decay (see Fig.\ \ref{fg:auger}), leads to the population of doubly ionized final states localized on the initially ionized unit \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415}.  The normal Auger decay process in clusters proceeds similarly to that in atoms or molecules. However, in the case of a core excited state, the resonant Auger process competes with the process of delocalization of the excited electron within the cluster. If the initially core excited electron delocalizes within the lifetime of the core hole, then normal instead of resonant Auger decay is observed \citep{Bjorneholm95:3017}. 
%
%
In a solution, the electronic decay processes initiated by x-ray photoabsorption are different compared to those in rare gas clusters due to the shorter distances and stronger interatomic interactions. In particular, the solvent molecules have two effects -- first, they strongly affect the excited \citep{miteva16:16671} or ionized states of the ion and second, they can participate in the decay processes, leading to the population of charge-separated final states, and ionization of the surrounding environment \citep{Pokapanich09:7264,Pokapanich11:13430,Stumpf16:237,ceolin17:263003,unger17:708}. Moreover, the process of delocalization of the initially excited electron also occurs in aqueous solutions \citep{Nordlund07:217406,Ottosson11:13489}. In the case of pure water, the delocalization of the O1s excited electron takes place on a femtosecond to sub-femtosecond time scale depending on the photon energy, thus being commensurate with the lifetime of the O1s core hole which is $\sim$3.6\,fs \citep{Nordlund07:217406}. The O K-edge is located in the soft x-ray range of photon energies, $\sim$533\,eV in water vapor \citep{wagner80:1445}. Going higher in photon energy, to the tender and hard x-ray regimes, the lifetimes of the core ionized or core excited states become even shorter, on the order of 1\,fs. And thus, it is even more imperative to reveal whether the delocalization of the core excited electron occurs within the lifetime of the core hole.

>>>>

It has been well understood in atoms and molecules in gas phase that the course of a decay cascade depends on the character of the initially populated states \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415,Miteva14:164303,travnikova16:213001,Gokhberg14:661,Trinter14:664}.
%The course of a decay cascade depends on the character of the initially populated states. This has been well understood in atoms and molecules in gas phase  \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415,Miteva14:164303,travnikova16:213001,Gokhberg14:661,Trinter14:664}. In the case of a core ionized state, the Auger decay process, designated as normal Auger decay (see Fig.\ \ref{fg:auger}), leads to the population of doubly ionized final states localized on the initially ionized unit \citep{stoychev08:074307,Demekhin08:043421,Demekhin09:104303,Ouchi11:053415}.  The normal Auger decay process in clusters proceeds similarly to that in atoms or molecules. However, in the case of a core excited state, the resonant Auger process competes with the process of delocalization of the excited electron within the cluster. If the initially core excited electron delocalizes within the lifetime of the core hole, then normal instead of resonant Auger decay is observed \citep{Bjorneholm95:3017}. 
%
%
In a solution, the electronic decay processes initiated by x-ray photoabsorption are different compared to those in rare gas clusters due to the shorter distances and stronger interatomic interactions. In particular, the solvent molecules have two effects -- first, they strongly affect the excited \citep{miteva16:16671} or ionized states of the ion and second, they can participate in the decay processes, leading to the population of charge-separated final states, and ionization of the surrounding environment \citep{Pokapanich09:7264,Pokapanich11:13430,Stumpf16:237,ceolin17:263003,unger17:708}. Moreover, the process of delocalization of the initially excited electron also occurs in aqueous solutions \citep{Nordlund07:217406,Ottosson11:13489} on a femtosecond to sub-femtosecond time scale.

%. In the case of pure water, the delocalization of the O1s excited electron takes place on a femtosecond to sub-femtosecond time scale depending on the photon energy, thus being commensurate with the lifetime of the O1s core hole which is $\sim$3.6\,fs \citep{Nordlund07:217406}. The O K-edge is located in the soft x-ray range of photon energies, $\sim$533\,eV in water vapor \citep{wagner80:1445}. 
Going higher in photon energy, to the tender and hard x-ray regimes, the lifetimes of the core ionized or core excited states become even shorter, on the order of 1\,fs. And thus, it is even more imperative to reveal whether the delocalization of the core excited electron occurs within the lifetime of the core hole.

### 2018-04-25 methods.tex
apr 25
jpcl draft

For the present experiment we used the newly operational microjet setup that was specifically designed for the HAXPES station of the GALAXIES beamline \citep{ceolin13:188,rueff15:175}. A differentially-pumped tube in which the microjet head is inserted, is mounted on a 3-axes motorized manipulator in front of the spectrometer lens. Two holes of 2\,mm diameter allow the photons to go in and out. At the end of the tube and in front of the lens, a 500\,$\mu$m diameter hole skimmer allows the electrons created at the interaction point to go in the direction of the spectrometer. The microjet head is mostly composed of a 30\,$\mu$m diameter vertical glass capillary facing a temperature-controlled catcher in CuBe having a 300\,$\mu$m hole, and a camera. Piezo motors allow their precise alignment relative to each other and to the photon beam. The catcher is placed at a distance of about 5\,mm from the capillary and is permanently pumped in order to extract the liquid. For the present experiment, a 0.5M KCl aqueous solution is injected in the capillary by a high performance liquid chromatography (HPLC) pump with a constant flux of 1.6\,ml/min. The alignment of the setup is performed on the KCl aqueous solution by measuring the water O1s x-ray photoelectron peak intensity and by optimizing the liquid vs gas phase ratio. The pressure in the main chamber is kept below the 10$^{-5}$\,mbar range whereas it is kept at about 10$^{-4}$\,mbar in the differentially-pumped tube when the HPLC pump is ON. Our equipment is an updated version of the equipment used in Ref.\ \cite{faubel88:269}. The aqueous potassium chloride solution was prepared by mixing >99\% KCl salt with deionized water. Filtering and degazing procedures were systematically performed before injecting the solution. The spectrometer resolution of about 0.6\,eV was achieved with the 500\,eV pass energy and 0.5\,mm slits. The photon energy resolution achieved at 2.8\,keV and 3.6\,keV was about 0.3\,eV and 0.4\,eV, respectively. The experimental 2D maps representing the evolution of the KLL Auger spectra in the vicinity of the \cli and \ki~K-edges, as a function of the photon energy, are shown in Figs.\ \ref{fg:2dmap_k} and \ref{fg:2dmap_cl}, respectively. The aqueous \ki~and \cli~1s ionization potentials were measured at h$\nu = 5$\,keV and calibrated on the liquid contribution of the O1s XPS spectrum \cite{winter06:1176}. The maps were also calibrated using the O1s photoelectron line of liquid water but at photon energies close to the potassium and chloride 1s ionization thresholds.

@@

The theoretical X-ray absorption spectra were computed for the hexa-coordinated clusters of both ions, \ki(H$_2$O)$_6$ and \cli(H$_2$O)$_6$, which can be considered as representatives of the complete first solvation shell of the two ions \citep{Ohtaki93:1157,soper06:180,ma14:1006}. The two structures shown in Fig.\ \ref{fg:xas_kcl} were optimized at the DFT level of theory using the B3LYP functional and the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,Blaudeau97:5016}. The geometry optimization was performed with the Gaussian 09 package \citep{g09}. In order to obtain a realistic structure for \ki~corresponding to the bulk solution, we carried out constrained geometry optimization by choosing the equilibrium gas-phase geometries \citep{lee99:3995,lee02:5509} belonging to the D$_3$ point group and then increasing the angle $\theta$ between the K-O bond and the $C_3$ axis to 55$^{\circ}$. This angle was chosen such that the O-K-O angles are around the maxima of the angular distributions obtained from quantum mechanics/molecular mechanics simulations in Ref.\ \citep{ma14:1006}. Moreover, we fixed the K-O distance to 2.840~\AA, such that it corresponds to the distances from other theoretical and experimental works \citep{Ohtaki93:1157,soper06:180,ma14:1006}.
%\citep{ge13:13169,gora00:7,Ohtaki93:1157,soper06:180,ma14:1006}.
%Moreover, we fixed the K-O and Cl-O distances to 2.840~\AA~and 3.140~\AA, respectively, such that they correspond to the distances from other theoretical and experimental works \citep{ge13:13169,gora00:7,Ohtaki93:1157,soper06:180,ma14:1006}.

The energies and transition moments of the core excited states of the bare ions and microsolvated clusters were computed with the algebraic diagrammatic construction method for the polarization propagator \citep{sch82:2395} within the core-valence separation approximation \citep{bar85:867,ced80:206,ced81:1038} (CVS-ADC(2)x) as implemented in the Q-Chem package \citep{Wenzel14:1900,Wenzel14:4583,Wormit14:774,QChem2015}. In the case of \cli~the 6-311++G(3df,3pd) basis set \citep{Krishnan80:650,McLean80:5639} (excluding f functions) was used on all atoms, whereas in the case of \ki~we used the 6-311+G(2d,p) basis set \citep{Krishnan80:650,Blaudeau97:5016} on all atoms, and two additional sets of s, p and d diffuse functions were added on K. The use of a smaller basis set in the case of K is due to the higher number of atomic orbitals compared to the case of Cl, and therefore, prohibitively high cost of the CVS-ADC(2)x computation. In our calculations the core space comprises the 1s orbital of K$^{+}$ or \cli, whereas the remaining occupied orbitals are included in the valence space. For the calculations of the XAS spectra we used the C$_2$ point group in the case of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$. To account for the lifetime broadening due to the Auger decay of the core excited states, we convolved the theoretical spectra with a Lorentzian function of FWHM 0.74\,eV and 0.62\,eV in the case of \ki~and \cli, respectively \citep{Krause79:329}. Additionally, we convolved the theoretical spectra with a Gaussian profile to also account for the experimental resolution (see Fig.\ \ref{fg:xas_kcl}). We analyzed the core excited states by expanding the natural orbitals occupied by the excited electron (singly occupied natural orbitals, SONOs) $\psi_{i}$ of the microsolvated clusters in the basis of SONOs of the bare K$^{+}$ or \cli~ions, $\chi_{nl}$
%
\begin{equation}\label{eq:sono_proj}
\psi_{i} = \sum_{nl} a^{i}_{nl} \chi_{nl}
\end{equation}
%
where $n$ and $l$ stand for the principal and orbital quantum numbers as described in Ref.\ \citep{miteva16:16671}. The expansion coefficients $a^{i}_{nl}$ show the degree of delocalization of the excited electron and the mixing of the core excited states in the ligand field created by the surrounding water molecules (see Fig.\ \ref{fg:xas_kcl}).

The final states following KLL resonant Auger decay of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$ were computed at the Configuration Interaction Singles (CIS) level using the Graphical Unitary Group Approach (GUGA) as implemented in the GAMESS-US package \citep{GUGA_PhysScr_21,GUGA_JCP_70,GUS}. In order to account for the relaxation effects upon core ionization, we employed a restricted open-shell Hartree-Fock reference wave function with a hole in the 2s orbital of both \ki~and \cli.  We used the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,McLean80:5639,Blaudeau97:5016} on all atoms. Additionally, the basis set was augmented with two sets of s, p, d diffuse functions in the case of \ki, and three sets of s, p, d diffuse functions in the case of \cli. The larger basis set employed in the case of Cl was necessary in order to ensure the convergence of the excited states. The active space comprises the 2s and 2p orbitals of K/Cl with occupancy fixed to 6 and all virtual orbitals with occupancy fixed to 1. The remaining doubly occupied orbitals were frozen in the calculation. \citep{mosnier16:061401}

>>>>

For the present experiment we used the newly operational microjet setup that was specifically designed for the HAXPES station of the GALAXIES beamline \citep{ceolin13:188,rueff15:175}. {\color{red}CAN WE MAKE THIS SHORTER? A differentially-pumped tube in which the microjet head is inserted, is mounted on a 3-axes motorized manipulator in front of the spectrometer lens. Two holes of 2\,mm diameter allow the photons to go in and out. At the end of the tube and in front of the lens, a 500\,$\mu$m diameter hole skimmer allows the electrons created at the interaction point to go in the direction of the spectrometer. The microjet head is mostly composed of a 30\,$\mu$m diameter vertical glass capillary facing a temperature-controlled catcher in CuBe having a 300\,$\mu$m hole, and a camera. Piezo motors allow their precise alignment relative to each other and to the photon beam. The catcher is placed at a distance of about 5\,mm from the capillary and is permanently pumped in order to extract the liquid. For the present experiment, a 0.5M KCl aqueous solution is injected in the capillary by a high performance liquid chromatography (HPLC) pump with a constant flux of 1.6\,ml/min. The alignment of the setup is performed on the KCl aqueous solution by measuring the water O1s x-ray photoelectron peak intensity and by optimizing the liquid vs gas phase ratio. The pressure in the main chamber is kept below the 10$^{-5}$\,mbar range whereas it is kept at about 10$^{-4}$\,mbar in the differentially-pumped tube when the HPLC pump is ON. Our equipment is an updated version of the equipment used in Ref.\ \cite{faubel88:269}.} The aqueous potassium chloride solution was prepared by mixing >99\% KCl salt with deionized water. Filtering and degazing procedures were systematically performed before injecting the solution. The spectrometer resolution of about 0.6\,eV was achieved with the 500\,eV pass energy and 0.5\,mm slits. The photon energy resolution achieved at 2.8\,keV and 3.6\,keV was about 0.3\,eV and 0.4\,eV, respectively. The experimental 2D maps representing the evolution of the KLL Auger spectra in the vicinity of the \cli and \ki~K-edges, as a function of the photon energy, are shown in Figs.\ \ref{fg:2dmap_k} and \ref{fg:2dmap_cl}, respectively. The aqueous \ki~and \cli~1s ionization potentials were measured at h$\nu = 5$\,keV and calibrated on the liquid contribution of the O1s XPS spectrum \cite{winter06:1176}. The maps were also calibrated using the O1s photoelectron line of liquid water but at photon energies close to the potassium and chloride 1s ionization thresholds.

@@

The theoretical X-ray absorption spectra were computed for the hexa-coordinated clusters of both ions, \ki(H$_2$O)$_6$ and \cli(H$_2$O)$_6$, which can be considered as representatives of the complete first solvation shell of the two ions \citep{Ohtaki93:1157,soper06:180,ma14:1006}. The two structures shown in Fig.\ \ref{fg:xas_kcl} were optimized at the DFT level of theory using the B3LYP functional and the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,Blaudeau97:5016}. The geometry optimization was performed with the Gaussian 09 package \citep{g09}. In order to obtain a realistic structure for \ki(H$_2$O)$_6$ corresponding to the bulk solution, we carried out constrained geometry optimization by fixing the K-O distance to 2.840~\AA~and increasing the angle $\theta$ between the K-O bond and the $C_3$ axis to 55$^{\circ}$ using the equilibrium gas-phase geometry belonging to the D$_3$ point group as a starting point \citep{lee99:3995,lee02:5509}.
%  and then increasing the angle $\theta$ between the K-O bond and the $C_3$ axis to 55$^{\circ}$. This angle was chosen such that the O-K-O angles are around the maxima of the angular distributions obtained from quantum mechanics/molecular mechanics simulations in Ref.\ \citep{ma14:1006}. Moreover, we fixed the K-O distance to 2.840~\AA, such that it corresponds to the distances from other theoretical and experimental works \citep{Ohtaki93:1157,soper06:180,ma14:1006}.
%\citep{ge13:13169,gora00:7,Ohtaki93:1157,soper06:180,ma14:1006}.
%Moreover, we fixed the K-O and Cl-O distances to 2.840~\AA~and 3.140~\AA, respectively, such that they correspond to the distances from other theoretical and experimental works \citep{ge13:13169,gora00:7,Ohtaki93:1157,soper06:180,ma14:1006}.

The energies and transition moments of the core excited states of the bare ions and microsolvated clusters were computed with the algebraic diagrammatic construction method for the polarization propagator \citep{sch82:2395} within the core-valence separation approximation \citep{bar85:867,ced80:206,ced81:1038} (CVS-ADC(2)x) as implemented in the Q-Chem package \citep{Wenzel14:1900,Wenzel14:4583,Wormit14:774,QChem2015}. In the case of \cli~the 6-311++G(3df,3pd) basis set \citep{Krishnan80:650,McLean80:5639} (excluding f functions) was used on all atoms, whereas in the case of \ki~we used the 6-311+G(2d,p) basis set \citep{Krishnan80:650,Blaudeau97:5016} on all atoms, and two additional sets of s, p and d diffuse functions were added on K. %The use of a smaller basis set in the case of K is due to the higher number of atomic orbitals compared to the case of Cl, and therefore, prohibitively high cost of the CVS-ADC(2)x computation. 
In our calculations the core space comprises the 1s orbital of K$^{+}$ or \cli, whereas the remaining occupied orbitals are included in the valence space. %For the calculations of the XAS spectra we used the C$_2$ point group in the case of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$. 
To account for the experimental resolution and the lifetime broadening due to the Auger decay of the core excited states, we convolved the theoretical spectra with a Gaussian profile and a Lorentzian function of FWHM 0.74\,eV and 0.62\,eV in the case of \ki~and \cli, respectively \citep{Krause79:329}. In order to understand the mixing of the core excited states in the ligand field created by the surrounding water molecules, we analyzed the core excited states of the hexa-coordinated clusters by expanding the natural orbitals occupied by the excited electron (singly occupied natural orbitals, SONOs) in the basis of SONOs of the bare K$^{+}$ or \cli~ions, as described in Ref.\ \citep{miteva16:16671}.

The final states following KLL resonant Auger decay of K$^{+}$(H$_2$O)$_6$ and \cli(H$_2$O)$_6$ were computed at the Configuration Interaction Singles (CIS) level using the Graphical Unitary Group Approach (GUGA) as implemented in the GAMESS-US package \citep{GUGA_PhysScr_21,GUGA_JCP_70,GUS}. In order to account for the relaxation effects upon core ionization, we employed a restricted open-shell Hartree-Fock reference wave function with a hole in the 2s orbital of both \ki~and \cli.  We used the 6-311++G(2d,2p) basis set \citep{Krishnan80:650,McLean80:5639,Blaudeau97:5016} on all atoms. Additionally, the basis set was augmented with two sets of s, p, d diffuse functions in the case of \ki, and three sets of s, p, d diffuse functions in the case of \cli. %The larger basis set employed in the case of Cl was necessary in order to ensure the convergence of the excited states. 
The active space comprises the 2s and 2p orbitals of K/Cl with occupancy fixed to 6 and all virtual orbitals with occupancy fixed to 1. The remaining doubly occupied orbitals were frozen in the calculation. \citep{mosnier16:061401}
